---
title: "whitebarkpine_esa"
format: html
editor: visual
---

# Exploring whitebark pine in FIA dataframe

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE


library(dplyr)
library(ggplot2)
library(readr)
library(tigris)
library(leaflet)
library(sf)
library(basemaps)
library(sf)
library(ggspatial)  
library(grid)    
library(patchwork)


theme_set(theme_bw())



```

# Whitebark pine for Montana, Idaho, and Wyoming

SPCD = 101 STATECD = 16, 30, 56

```{r}

whitebarkpine_all <- read_csv("data_processed/final_whitebark_pine.csv")



whitebarkpine_all$PLT_CN <- as.character(whitebarkpine_all$PLT_CN)
whitebarkpine_all$TREE_CN <- as.character(whitebarkpine_all$TREE_CN)
whitebarkpine_all$COND_CN <- as.character(whitebarkpine_all$COND_CN)


knitr::kable(head(whitebarkpine_all))


```

**We have a total of 28,385 identified whitebark pine trees from the annual inventory period from MT, ID, WY**

## How many trees have been surveyed how many times in each state?

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

whitebark_pine_measures <- whitebarkpine_all |> 
  group_by(STATECD, TREE_COMPOSITE_ID) |>
  summarize(n = length(unique(INVYR))) 


knitr::kable(whitebark_pine_measures |>
               group_by(STATECD, n) |>
               tally() |>
               rename(n_measures = n,
                      n_trees = nn))


```

**Notably, zero trees in Wyoming (STATECD = 56) have been surveyed more than one time.**

## Filter dataframe to only include remeasured plots so we can look at mortality

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE


whitebark_pine_repeats <- filter(whitebark_pine_measures, n > 1) |>
  left_join(whitebarkpine_all)


whitebark_pine_repeats <- whitebark_pine_repeats %>% 
  relocate(STATECD, TPA_UNADJ, TREE_COMPOSITE_ID, INVYR, CYCLE, MEASYEAR, STATUSCD, DIA)

length(unique(whitebark_pine_repeats$TREE_COMPOSITE_ID))



```

This filters our search down to 8,147 trees that can be tracked from the annual inventory period through time (check back in on this when FIAdb has been updated - most recent update occurred June 28th, 2024 - talk to Renata about updating the database)

# Identify the trees that were alive at one point and then died

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
#| 
# Filter trees that died
dead_trees <- whitebark_pine_repeats %>%
  group_by(TREE_COMPOSITE_ID) %>%
  filter(any(STATUSCD == 2) & any(STATUSCD == 1)) %>%
  ungroup()

# Extract the DIA values when trees were alive so we can plot death as a function of the state variable (size)
alive_dia <- dead_trees %>%
  filter(STATUSCD == 1) %>%
  select(TREE_COMPOSITE_ID, DIA)

# Filter trees that survived - last status is alive 
survived_trees <- whitebark_pine_repeats %>%
  group_by(TREE_COMPOSITE_ID) %>%
  filter(STATUSCD == 1) %>%
  ungroup()

length(unique(alive_dia$TREE_COMPOSITE_ID)) # 1,904 trees died
length(unique(survived_trees$TREE_COMPOSITE_ID)) # 4,960 trees survived

# Find the mean DIA of all the remeasured trees
mean_dia_all <- mean(whitebark_pine_repeats$DIA, na.rm = TRUE)
mean_dead_trees <- mean(alive_dia$DIA)

# Find the max INVY
max(whitebark_pine_repeats$INVYR)
min(whitebark_pine_repeats$INVYR)

# Plot histogram 
ggplot() +
  geom_histogram(data = survived_trees, aes(x = DIA, fill = "Survived"), binwidth = 1, color = "white", alpha = 0.5, position = "identity") + 
  geom_histogram(data = alive_dia, aes(x = DIA, fill = "Died"), binwidth = 1, color = "white", alpha = 0.6, position = "identity") +
  labs(title = "Whitebark pine mortality and survival of remeasured trees from 2003-2019",
       x = "Diameter (in.)",
       y = "Number of trees", 
       fill = "Tree Status") +
  geom_vline(aes(xintercept = mean_dead_trees), color = "red", linetype = "dashed") + 
  geom_vline(aes(xintercept = mean_dia_all), color = "black", linetype = "dashed") + 
  theme_minimal() + 
  scale_fill_manual(values = c("Died" = "#006699", "Survived" = "#9999FF"))

##need to incorporate expansion factor

```

So 1,904 whitebark pine trees that were alive had died from the annual inventory cycle from MT and ID (note - does not include Wyoming)

Mean DIA of dead trees is 8.922 in Mean DIA of all the trees is 8.767 in The mean for value for the trees that died is higher than the mean value for all the trees pooled that lived and died

27.7% of whitebark pine trees died from 2003-2019

Need to include expansion factor for trees on microplots to scale up to the plot level 

# Making Fitts (see: Fitts et al. 2021) type tables for individual trees, mortality agents vs disturbance codes for whitebark pine --> compounded disturbances vs. simple disturbances 

## First we have to deal with the expansion factor on microplots which applies to individuals with an expansion factor of 74.965282

```{r}

whitebark_pine_microplot <- whitebark_pine_repeats %>% 
  filter((TPA_UNADJ == 74.965282) | 
   (is.na(TPA_UNADJ) & (lag(TPA_UNADJ) == 74.965282 | lead(TPA_UNADJ) == 74.965282))
  ) 

# okay now add the expansion factor to scale microplot trees up to the subplot level so we can compare total number of tree mortalitt
whitebark_pine_microplot <- whitebark_pine_microplot %>%
  mutate(microplot_corrected = 1 * 12.4567479)

whitebark_pine_repeats <- whitebark_pine_repeats %>%
  left_join(select(whitebark_pine_microplot, TREE_COMPOSITE_ID, microplot_corrected), by = "TREE_COMPOSITE_ID")

str(whitebark_pine_repeats)
  
# Plot histogram of live vs dead
# Filter trees that died
dead_trees_corrected <- whitebark_pine_repeats %>%
  group_by(TREE_COMPOSITE_ID) %>%
  filter(any(STATUSCD == 2) & any(STATUSCD == 1)) %>%
  ungroup()


# Extract the DIA values when trees were alive so we can plot death as a function of the state variable (size)
alive_dia_corrected <- dead_trees_corrected %>%
  filter(STATUSCD == 1) %>%
  select(TREE_COMPOSITE_ID, DIA, microplot_corrected) %>%
  mutate(weight = if_else(is.na(microplot_corrected), 1, microplot_corrected))

# Filter trees that survived - last status is alive 
survived_trees_corrected <- whitebark_pine_repeats %>%
  group_by(TREE_COMPOSITE_ID) %>%
  filter(STATUSCD == 1) %>%
  ungroup() %>%
  select(TREE_COMPOSITE_ID, DIA, microplot_corrected) %>%
  mutate(weight = if_else(is.na(microplot_corrected), 1, microplot_corrected))

# Calculate mean DIA
mean_dia_all <- mean(whitebark_pine_repeats$DIA, na.rm = TRUE)
mean_dead_trees <- mean(alive_dia_corrected$DIA, na.rm = TRUE)

#plot percentages of morality on histogram
# Calculate frequencies and percentages

bins <- seq(0, 30, by = 1)

alive_dia_binned <- alive_dia_corrected %>%
  mutate(bin = cut(DIA, breaks = bins, include.lowest = TRUE)) %>%
  group_by(bin) %>%
  summarise(count = sum(weight))

survived_trees_binned <- survived_trees_corrected %>%
  mutate(bin = cut(DIA, breaks = bins, include.lowest = TRUE)) %>%
  group_by(bin) %>%
  summarise(count = sum(weight))

# Combine data and calculate percentages
combined_binned <- full_join(alive_dia_binned, survived_trees_binned, by = "bin", suffix = c("_died", "_survived")) %>%
  replace_na(list(count_died = 0, count_survived = 0)) %>%
  mutate(total = count_died + count_survived,
         perc_died = (count_died / total) * 100,
         perc_survived = (count_survived / total) * 100)


# Plot histogram 
mortality_hist <-ggplot() +
  geom_histogram(data = survived_trees_corrected, aes(x = DIA, weight = weight, fill = "Survived"), binwidth = 1, color = "white", alpha = 0.75, position = "identity") + 
  geom_histogram(data = alive_dia_corrected, aes(x = DIA, weight = weight, fill = "Died"), binwidth = 1, color = "white", alpha = 0.6, position = "identity") +
  labs(title = "Whitebark Pine Mortality and Survival of remeasured trees from 2003-2019 in the interior west",
       x = "Diameter (in.)",
       y = "Number of Trees", 
       fill = "Tree Status") +
  geom_vline(aes(xintercept = mean_dead_trees), color = "red", linetype = "dashed") + 
  geom_vline(aes(xintercept = mean_dia_all), color = "black", linetype = "dashed") + 
  theme_bw() + 
  scale_fill_manual(values = c("Died" = "#006699", "Survived" = "#9999FF"))

ggsave("mortality_hist.png", mortality_hist, width = 10, height = 6)

##okay this table is adjusted for saplings on the microplot


```

# Okay now explore some the disturbance agents and damage agent codes

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE


whitebark_pine_agentcd <- whitebark_pine_repeats %>%
  filter(STATUSCD == 2)

print(unique(whitebark_pine_agentcd$AGENTCD)) # 10, 20, 30, 40, 50, 60, 70
print(length(unique(whitebark_pine_agentcd$TREE_COMPOSITE_ID))) #5091

whitebark_pine_agentcd_unique <- whitebark_pine_agentcd %>%
  group_by(TREE_COMPOSITE_ID) %>%
  arrange(TREE_COMPOSITE_ID, INVYR) %>%
  slice(1) %>%
  ungroup()

whitebark_pine_agentcd_unique$AGENTCD <- factor(whitebark_pine_agentcd_unique$AGENTCD, exclude = NULL)

mortality_labels <- c(
  "10" = "Insect",
  "20" = "Disease",
  "30" = "Fire",
  "40" = "Animal",
  "50" = "Weather",
  "60" = "Vegetation",
  "70" = "Unknown",
  "NA" = "NA"
)

whitebark_pine_agentcd_unique$AGENTCD <- as.character(whitebark_pine_agentcd_unique$AGENTCD)

# Create a bar plot showing the frequency distribution of AGENTCD
agent_hist <- ggplot(data = whitebark_pine_agentcd_unique, aes(x =  as.factor(AGENTCD), fill = AGENTCD)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +
  labs(title = "Mortality Causing Agents of Whitebark Pine",
       x = "Mortality Agent",
       y = "Frequency") +
  scale_x_discrete(labels = mortality_labels) +
  scale_fill_manual(values = c("10" = "black",
                               "20" = "darkorange",
                               "30" = "darkred",
                               "40" = "brown",
                               "50" = "darkblue",
                               "60" = "darkgreen",
                               "70" = "purple4",
                               "NA" = "gray")) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ylim(0, 2800)

ggsave("agentcd_hist.png", agent_hist, width = 10, height = 6)


```

# Okay now add condition level disturbance codes to the mix

```{r}

#| echo: FALSE
#| message: FALSE
#| warning: FALSE

whitebark_pine_agentcd_unique$DSTRBCD1 <- as.factor(whitebark_pine_agentcd_unique$DSTRBCD1)
whitebark_pine_agentcd_unique$DSTRBCD2 <- as.factor(whitebark_pine_agentcd_unique$DSTRBCD2)
whitebark_pine_agentcd_unique$DSTRBCD3 <- as.factor(whitebark_pine_agentcd_unique$DSTRBCD3)

print(unique(whitebark_pine_agentcd$DSTRBCD1)) # 0, 21, 22, 12, 31, 30, 32, 20, 95, 50, 10, 46, 92, 60, 52

# Define the disturbance code labels
disturbance_labels <- c(
  "0" = "No visible disturbance",
  "10" = "Insect damage",
  "12" = "Insect damage to trees, including seedlings and saplings",
  "20" = "Disease damage",
  "21" = "Disease damage to understory vegetation",
  "22" = "Disease damage to trees, including seedlings and saplings",
  "30" = "Fire damage (from crown and ground fire, either prescribed or natural)",
  "31" = "Ground fire damage",
  "32" = "Crown fire damage",
  "40" = "Animal damage",
  "41" = "Beaver (includes flooding caused by beaver)",
  "42" = "Porcupine",
  "43" = "Deer/ungulate",
  "44" = "Bear",
  "45" = "Rabbit",
  "46" = "Domestic animal/livestock (includes grazing)",
  "50" = "Weather damage",
  "51" = "Ice",
  "52" = "Wind (includes hurricane, tornado)",
  "53" = "Flooding (weather induced)",
  "54" = "Drought",
  "60" = "Vegetation (suppression, competition, vines)",
  "70" = "Unknown / not sure / other",
  "80" = "Human-induced damage",
  "90" = "Geologic disturbances",
  "91" = "Landslide",
  "92" = "Avalanche track", 
  "NA" = "NA"
)

ggplot(data = whitebark_pine_agentcd_unique, aes(x = DSTRBCD1)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +
  labs(title = "Disturbance Agents to the forest stand",
       x = "Disturbance Code",
       y = "Frequency") +
  scale_x_discrete(labels = disturbance_labels) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# pooling some of the forest disturbance agents 
whitebark_pine_agentcd_unique <- whitebark_pine_agentcd_unique %>%
  mutate(PooledCategory = case_when(
    DSTRBCD1 == 0  ~ "No visible damage",
    DSTRBCD1 %in% c("10", "12") ~ "Insect damage",
    DSTRBCD1 %in% c("20", "21", "22") ~ "Disease damage",
    DSTRBCD1 %in% c("30", "31", "32") ~ "Fire damage",
    DSTRBCD1 %in% c("46", "50", "52", "60", "92", "95") ~ "Other",
    TRUE ~ as.character(DSTRBCD1)  # Retain other codes as they are
  ))

whitebark_pine_agentcd_unique$PooledCategory <- factor(whitebark_pine_agentcd_unique$PooledCategory)

ggplot(data = whitebark_pine_agentcd_unique, aes(x = PooledCategory)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +
  labs(title = "Distribution of Disturbance Codes",
       x = "Disturbance Category",
       y = "Frequency") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
# Make a joint plot with mortality and disturbance
```{r}
#  custom colors for the disturbance categories
custom_colors <- c(
  "Disease damage" = "darkorange",
  "Fire damage" = "darkred",
  "Insect damage" = "black",
  "No visible damage" = "darkblue",
  "Other" = "purple4"
)

# plot this histogram
dist_agent_hist <- ggplot(data = whitebark_pine_agentcd_unique, aes(x = AGENTCD, fill = PooledCategory)) +
  geom_bar(position = "dodge", color = "white") +
  labs(title = "Disturbance categories of whitebark pine by mortality agent",
       x = "Mortality Agent",
       y = "Number of trees",
       fill = "Disturbance Category") +
  scale_x_discrete(labels = mortality_labels) +
  scale_fill_manual(values = custom_colors) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


ggsave("dist_agent.png", dist_agent_hist, width = 10, height = 6)


```

# Now we are exploring how well crossdated the tree ring data are

```{r}

#| echo: FALSE
#| message: FALSE
#| warning: FALSE

# reading in the wbp range shapefile
wbp_range_data <- sf::st_read("WBP_Range_2014_v11/WBP_range_2014_d.shp") 
wbp_range_data_transformed <- st_transform(wbp_range_data, crs = st_crs(3857))

# read in tree ring occurrence data
tree_ring_dat <- read_csv("data_processed/wbp_fiadb.csv")
head(tree_ring_dat)

#only keep trees where verified = y
tree_ring_usable <- tree_ring_dat %>% 
  filter(VERIFY == "y")

knitr::kable(as.data.frame(table(tree_ring_usable$STATECD.x)), 
      col.names = c("STATECD.x", "Count"), 
      caption = "Number of Samples for Each State Code")


```

